Species-specific effects of the Urban Heat Island on tree growth across Berlin

### 2020-03-31 analysis/paper/paper.Rmd
updated paper

where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$. 
Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coeffecient:

@@

Nested data structures (e.g. due to similar road [type]) can be accounted for by introducing random effects [@wood2017], while spatial dependence between observations can be included by constructing smoothing functions with e.g. northings and eastings, as for example done in [@augustin2009].
Ultimately, the implementation of a such a GAMM will alow for establishing continuous prediction surfaces of growth potential (approximated via $DBH$) for individual species across urban areas (including parks) of Berlin.

Currently, $DBH$ has been modelled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020 (see Section$~$\@ref(sec:results)).
The general form of this model is:

\begin{equation}
Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
(\#eq:lme)
\end{equation}
where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
note, that for computational efficiency each genera was modelled seperately.
Further, models were established for the three most abundant species per genera with at least 1000 individuals.

@@

caption_trees <- "Binned age-distribution for trees in Berlin dataset, and entries missing age information."
caption_trees_short <- "Binned age-distribution for trees in Berlin dataset."

@@

**Note, that results below are preliminary and should be considered as a template for future outputs, rather than used for inference.**
The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increasing UHI loading,
while the most frequent genera, *Tilia*, features contrasting relationships between species.
The estimated effect sizes presented here are linear.
However, temperature may excert a non-linear control on absolute growth and, hence, applying a  method able to capture such dynamics may result in somewhat different effect sizes / behavior.
Additionally, if temperatures increase in the future under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure (i.e. using GAMM over linear models-).

@@

- validating the database with independent observations
- incorporating more pertinent covariates as dependent variables in the linear mixed model  
- testing multiple model structures with formal model selection procedures
- checking model resiudals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species. Note, that  complications presented by 'out-of-sample' predictions will be addressed.
  * assess potential age-dependent UHI impacts on individual species.  
  * repeating the above (GAMM) with total and incremental basal area as responses

>>>>

where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$. 
Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coefficient:

@@

Nested data structures (e.g. due to similar road [type]) can be accounted for by introducing random effects [@wood2017], while spatial dependence between observations can be included by constructing smoothing functions with e.g. northings and eastings, as for example done in [@augustin2009].
Ultimately, the implementation of a such a GAMM will allow for establishing continuous prediction surfaces of growth potential (approximated via $DBH$) for individual species across urban areas (including parks) of Berlin.

Currently, $DBH$ has been modeled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020 (see Section$~$\@ref(sec:results)).
The general form of this model is:

\begin{equation}
Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
(\#eq:lme)
\end{equation}
where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
note, that for computational efficiency each genera was modeled separately.
Further, models were established for the three most abundant species per genera with at least 1000 individuals.

@@

caption_trees <- "Binned age-distribution for trees in Berlin data set, and entries missing age information."
caption_trees_short <- "Binned age-distribution for trees in Berlin data set."

@@

**Note, that results below are preliminary and should be considered as a template for future outputs, rather than used for inference.**
The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increasing UHI loading,
while the most frequent genera, *Tilia*, features contrasting relationships between species.
The estimated effect sizes presented here are linear.
However, temperature may exert a non-linear control on absolute growth and, hence, applying a  method able to capture such dynamics may result in somewhat different effect sizes / behavior.
Additionally, if temperatures increase in the future under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure (i.e. using GAMM over linear models-).

@@

- validating the database with independent observations
- incorporating more pertinent covariates as dependent variables in the linear mixed model  
- testing multiple model structures with formal model selection procedures
- checking model residuals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species. Note, that  complications presented by 'out-of-sample' predictions will be addressed.
  * assess potential age-dependent UHI impacts on individual species.  
  * repeating the above (GAMM) with total and incremental basal area as responses
